---
title: Data munging
author: "Abhijit Dasgupta"
date: "March 25-27, 2019"
output: 
  xaringan::moon_reader:
    css: [default, './robot.css', './robot-fonts.css']
    #css: [default, metropolis, metropolis-fonts]
    nature:
      ratio: '16:9'
      highlightLanguage: R
      countIncrementalSlides: false
      highlightStyle: zenburn
      highlightLines: true
      slideNumberFormat: "%current%"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

layout: true

<div class="my-header">
<span>PS 312, March 2019</span></div>

---
class: middle, center

# Tidy data

---

# Make the computer happy

Just as we want code to make humans happy, we want data to make computers happy

Tidy data is a principle promoted by Dr. Hadley Wickham to make computers happy

--

The properties of a tidy dataset are:

1. Each variable forms a column
1. Each observation forms a row
1. Each type of observational unit forms a table. 

This forms a standardized way to structure a dataset, and so makes it easy for the 
analyst to develop standard pipelines. 

---
class: center, middle

## A tidy dataset is tidy in one way, but a messy dataset can be messy in many ways

<span style='text-align:right;font-size:30pt;'>Hadley Wickham</span>

---

# Messy data

A dataset can be messy in many many ways. Many of the more common issues are listed below:

- Column names contain values, not just variable names
- Multiple variables are stored in one column
- Variables are stored in both rows and columns
- Multiple types of observational types are stored in the same table
- A single observational unit is stored in multiple tables

Sometimes the messier format is better for data entry, but bad for data analyses. 


---

# Variables in column names

```{r}
library(tidyverse)
pew <- import('data/pew.csv')
head(pew, 4)
```

- This dataset has actual data in the column headers, rather than variable names. 
- We should ideally have 3 columns in this dataset: religion, income and frequency. 
- We can achieve this using a function called `gather` which takes a wide dataset and makes it tall. 

---

.left-column[
- Gather all the columns into two columns, `income` and `frequency`, by stacking the columns
- Don't include the variable `religion`
]
.right-column[
```{r, eval=T}
pew %>% 
  gather(income, frequency, -religion)
```
]

---

.pull-left[
```{r}
head(pew)
```
]
.pull-right[
```{r}
pew %>% 
  gather(income, frequency, -religion) %>% head(20)
```

]

---

# Multiple variables in column names

```{r}
tb <- import('data/tb.csv') %>% as_tibble()
head(tb)
```

- column headers include both sex and age

---

```{r}
tb %>% 
  gather(sex_age, n, -iso2, -year, -fu)


```

---

```{r}
tb %>% 
  gather(sex_age, n, -iso2, -year, -fu, na.rm=T)


```

---

```{r}
tb %>% 
  gather(sex_age, n, -iso2, -year, -fu, na.rm=T) %>% 
  separate(sex_age, c('sex','age'), sep=1) # by position

```

This still needs to be cleaned

---

# Variables stored in rows and columns

```{r}
weather <- import('data/weather.csv') %>% as_tibble()
weather
```

---

```{r}
weather %>% 
  gather(day, temp, d1:d31) #<<
```

`d1:d31` denotes all the variables physically between `d1` and `d31` in the dataset

> See what happens when you type `1:10` in the console

---

```{r}
weather %>% 
  gather(date, temp, d1:d31) %>% 
  spread(element, temp) #<<
```

---
class: middle, center

# Data cleaning

---



