---
title: "Data visualization"
author: "Abhijit Dasgupta"
date: "March 25-27, 2019"
output: 
  xaringan::moon_reader:
    css: [default, './robot.css', './robot-fonts.css']
    #css: [default, metropolis, metropolis-fonts]
    nature:
      ratio: '16:9'
      highlightLanguage: R
      countIncrementalSlides: false
      highlightStyle: docco
      highlightLines: true
      slideNumberFormat: "%current%"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "",
                      message = FALSE, warning = FALSE,
                      out.height="400px")
library(rio)
library(tidyverse)

```

layout: true

<div class="my-header">
<span>PS 312, March 2019</span></div>

---
class: middle, center
# Starting with ggplot2
---
# Why ggplot2?

We're making the decision to use `ggplot2` for graphics

+ Makes pretty good formatting choices out of the box
+ Works like pipes!!
+ Is declarative (tell it what you want) without getting caught up in minutae
+ Strongly leverages data frames (good practice)
+ Fast enough
+ There are good templates if you want to change the look
---
# Introduction to ggplot2


The `ggplot2` package is a very flexible and (to me) intuitive way of visualizing data.
It is based on the concept of layering elements on a canvas.

> This idea of layering graphics on a canvas is, to me, a nice way of building graphs

---
# Introduction to ggplot2

You need:

+ A `data.frame` object
+ _Aesthetic mappings_ (aes) to say what data is used for what purpose in the viz
    + x- and y-direction
    + shapes, colors, lines
+ A _geometry object_ (geom) to say what to draw
    + You can "layer" geoms on each other to build plots
    
---
# Introduction to ggplot2

`ggplot` used pipes before pipes were a thing. 

However, it uses the `+` symbol for
piping rather than the `%>%` operator, since it pre-dates the `tidyverse`

---
# Introduction to ggplot2

.pull-left[
```{r, eval=F}
library(ggplot2)
theme_set(theme_bw())
ggplot(mtcars, aes(x = wt, y = mpg)) + geom_point()
```

+ A `data.frame` object: mtcars
+ Aesthetic mapping: 
    - x-axis: wt
    - y-axis: mpg
+ Geometry:
    + geom_point: draw points
]
.pull-right[

```{r, eval=T, echo=F}
library(ggplot2)
ggplot(mtcars, aes(x = wt, y = mpg)) + geom_point()
```
]

---
# Introduction to ggplot2

.pull-left[
```{r, eval=F}
ggplot(mtcars, aes(x = wt, y = mpg)) + 
  geom_point()+ 
  geom_smooth()
```

+ A `data.frame` object: mtcars
+ Aesthetic mapping: 
    - x-axis: wt
    - y-axis: mpg
+ Geometry:
    + geom_point: draw points
    + geom_smooth: Add a layer which draws a best-fitting line
]
.pull-right[
```{r, eval=T, echo=F}
ggplot(mtcars, aes(x = wt, y = mpg)) + 
  geom_point()+ 
  geom_smooth()
```
]

---
# Introduction to ggplot2

.pull-left[
```{r, eval=F}
ggplot(mtcars, aes(x = wt, y = mpg)) + 
  geom_point()+ 
  geom_smooth()+
  labs(x = 'Weight', 
       y = 'Miles per gallon', 
       title = 'Fuel efficiency by weight',
       caption = "Source: mtcars dataset")
```

+ A `data.frame` object: mtcars
+ Aesthetic mapping: 
    - x-axis: wt
    - y-axis: mpg
+ Geometry:
    + geom_point: draw points
    + geom_smooth: Add a layer which draws a best-fitting line
]
.pull-right[
```{r, eval=T, echo=F}
ggplot(mtcars, aes(x = wt, y = mpg)) + 
  geom_point()+ 
  geom_smooth()+
  labs(x = 'Weight', 
       y = 'Miles per gallon', 
       title = 'Fuel efficiency by weight',
       caption = "Source: mtcars dataset")
```

]

---
class: middle, center
# Basic plots

---
# Histograms

```{r}
dos <- import('Data/FSI/Department of State.csv')
dos %>% 
  ggplot(aes(x = Award_Transaction_Value)) + geom_histogram()
```

---
# Histograms

```{r}
dos %>% 
  ggplot(aes(x = Award_Transaction_Value)) + geom_histogram()+
  scale_x_log10() # x-axis on log scale
```

---
# Density plots
```{r}
dos %>% 
  ggplot(aes(x = Award_Transaction_Value)) + geom_density()+
  scale_x_log10()

```

---
# Bar plots

```{r}
library(lubridate)
dos %>% 
  group_by(year = year(as_date(Award_Start_Date))) %>% 
  summarize(amount = sum(Award_Transaction_Value)) %>% 
  ggplot(aes(x = year, y = amount)) + # Note change in pipe operator
    geom_bar(stat='identity')
```

---
## Exercise

Using the `mtcars` dataset in R, create:

1. A histogram of the fuel efficiences (`mpg`) in the data set
2. A bar plot of frequencies of number of cylinders (`cyl`) in the car

---
## Solution

.pull-left[
```{r, fig.size=3}
ggplot(mtcars, aes(x = mpg)) + geom_histogram(binwidth=3)
# ggplot(mtcars) + geom_histogram(aes(x = mpg), binwidth = 3)
```
]
.pull-right[
```{r}
ggplot(mtcars, aes(x = factor(cyl))) + geom_bar()
```
]

---
class: middle, center

# Two continuous variables
---

# Scatter plot with best-fitting line

```{r}
ggplot(mtcars, aes(x = hp, y = mpg))+
  geom_point()+
  geom_smooth(method = 'lm')
```

---

## Line plot (for time series)
```{r, warning=F, message=FALSE}
library(forecast)
d <- data.frame(x = 1:length(gas), y = gas) # Australian monthly gas production
ggplot(d, aes(x, y)) + geom_line()
```
---
# Exercise

1. Create a scatter plot of sepal length and sepal width from the `iris` dataset, and 
add a smooth line through it

---

## Solution

```{r}
ggplot(iris, aes(Sepal.Length, Sepal.Width)) + geom_point() + geom_smooth()
```

---
class: center, middle

# Continuous variable with discrete variable

---

# Box plots

```{r}
dos %>% 
ggplot(aes(x = factor(year(as_date(Award_Start_Date))),
                y = Award_Transaction_Value))+
  geom_boxplot() +
  scale_y_log10()+
  labs(x = 'Year')
```

---
# Violin plots

```{r}
dos %>% 
ggplot(aes(x = factor(year(as_date(Award_Start_Date))),
                y = Award_Transaction_Value))+
  geom_violin() +
  scale_y_log10()+
  labs(x = 'Year')
```

---
## Exercise

1. Plot a boxplot of petal length by species using the `iris` dataset
---
## Solution

```{r}
ggplot(iris, aes(x = Species, y = Petal.Length))+geom_boxplot()
```

---
class: center, middle

# Grouped visualization

---

```{r}
grp_data <- dos %>% 
  group_by(Recipient_Location, year = year(as_date(Award_Start_Date))) %>% 
  summarize(amt = sum(Award_Transaction_Value)) %>% 
  filter(str_detect(Recipient_Location, '^C'))
ggplot(grp_data, aes(x = year, y = amt, color=Recipient_Location))+
  geom_line()+
  scale_y_log10()
```

---
```{r}
ggplot(grp_data, aes(x = year,  y = amt))+
  geom_line()+
  scale_y_log10()+
  facet_wrap(~Recipient_Location)
```

---
# Adding more strata

.pull-left[
```{r, eval=F}
dos %>% filter(str_detect(Recipient_Location, '^C')) %>%
ggplot(aes(x = year(as_date(Award_Start_Date)), 
           y = Award_Transaction_Value, 
           color = Recipient_Location, 
           shape = Award_Transaction_Type))+
  geom_point()+ 
  labs(x = 'Year', color='Location')+
  scale_y_log10()

```
]
.pull-right[
```{r, echo=F}
dos %>% filter(str_detect(Recipient_Location, '^C')) %>%
ggplot(aes(x = year(as_date(Award_Start_Date)), 
           y = Award_Transaction_Value, 
           color = Recipient_Location, 
           shape = Award_Transaction_Type))+
  geom_point()+ 
  labs(x = 'Year', color='Location')+
  scale_y_log10()

```
]
---
# Adding more strata

.pull-left[
```{r, eval=F}
dos %>% filter(str_detect(Recipient_Location, '^C')) %>%
ggplot(aes(x = year(as_date(Award_Start_Date)), 
           y = Award_Transaction_Value, 
           color = Recipient_Location, 
           shape = Award_Transaction_Type))+
  geom_jitter()+ 
  labs(x = 'Year', color='Location')+
  scale_y_log10()

```
]
.pull-right[
```{r, eval=T, echo=F}
dos %>% filter(str_detect(Recipient_Location, '^C')) %>%
ggplot(aes(x = year(as_date(Award_Start_Date)), 
           y = Award_Transaction_Value, 
           color = Recipient_Location, 
           shape = Award_Transaction_Type))+
  geom_jitter()+ 
  labs(x = 'Year', color='Location')+
  scale_y_log10()

```
]

---

# Error bars

```{r}
schools <- rio::import('Data/FSI/schools.rds')
schools %>% filter(tophead=='Elementary schools', 
                   head2=="Average hours in school day") %>% 
  filter(!is.na(State), State != 'United States') %>% 
  ggplot(aes(x = State, y = stats, ymin = stats - 2*se, 
             ymax = stats + 2*se)) +
  geom_pointrange()+
  labs(y = 'Avg hours in school day')+
  theme_bw()+
  theme(axis.text.x = element_text(angle=45, hjust = 1))
```


---
class: center, middle

# Maps

---


# Chloropleth

```{r, echo = F, out.height="800px", fig.align='center'}
library(choroplethr)
data(df_pop_county)
county_choropleth(df_pop_county,
                  title="US 2012 County Population Estimates",
                  legend = "Population")
```

---

# Using shapefiles
```{r, eval =T}
library(sf)
hrr_info <- st_read('~/Downloads/hrr_bdry-1/HRR_Bdry.SHP')
head(hrr_info)
# ggplot(hrr_info)+geom_sf()
# ggsave('map.png')
```

---


# Putting together multiple graphs

.pull-left[
```{r, echo = T, eval=F, fig.align = 'center'}
# install.packages('cowplot')
library(cowplot)
p1 <- ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
  geom_point() + facet_grid(. ~ Species) + stat_smooth(method = "lm") +
  background_grid(major = 'y', minor = "none") +
  panel_border() + theme(legend.position = "none")

# plot B
p2 <- ggplot(iris, aes(Sepal.Length, fill = Species)) +
  geom_density(alpha = .7) + theme(legend.justification = "top")
p2a <- p2 + theme(legend.position = "none")

# plot C
p3 <- ggplot(iris, aes(Sepal.Width, fill = Species)) +
  geom_density(alpha = .7) + theme(legend.position = "none")

# legend
legend <- get_legend(p2)

# align all plots vertically
plots <- align_plots(p1, p2a, p3, align = 'v', axis = 'l')

# put together bottom row and then everything
bottom_row <- plot_grid(plots[[2]], plots[[3]], legend, labels = c("B", "C"), rel_widths = c(1, 1, .3), nrow = 1)
plot_grid(plots[[1]], bottom_row, labels = c("A"), ncol = 1)

```
]
.pull-right[
```{r, echo = F, eval=T, fig.align = 'center'}
# install.packages('cowplot')
library(cowplot)
p1 <- ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
  geom_point() + facet_grid(. ~ Species) + stat_smooth(method = "lm") +
  background_grid(major = 'y', minor = "none") +
  panel_border() + theme(legend.position = "none")

# plot B
p2 <- ggplot(iris, aes(Sepal.Length, fill = Species)) +
  geom_density(alpha = .7) + theme(legend.justification = "top")
p2a <- p2 + theme(legend.position = "none")

# plot C
p3 <- ggplot(iris, aes(Sepal.Width, fill = Species)) +
  geom_density(alpha = .7) + theme(legend.position = "none")

# legend
legend <- get_legend(p2)

# align all plots vertically
plots <- align_plots(p1, p2a, p3, align = 'v', axis = 'l')

# put together bottom row and then everything
bottom_row <- plot_grid(plots[[2]], plots[[3]], legend, labels = c("B", "C"), rel_widths = c(1, 1, .3), nrow = 1)
plot_grid(plots[[1]], bottom_row, labels = c("A"), ncol = 1)

```
]

---
class: middle, center

# Interactive graphs

---

# Interactive graphs

There are several packages available in R to do interactive graphs. 

- plotly: A general plotting library from the folks at Plot.ly
- highchart: A port of the highcharter Javascript library
- dygraphs: A port of the dygraph Javascript library
- leaflet: Interactive maps

There are others as well. See [http://www.htmlwidgets.org](http://www.htmlwidgets.org])
for more details

---

# Plotly

.pull-left[
```{r,  eval=F, fig.align='center', out.height="400px"}
library(ggplot2)
library(plotly)
p=ggplot(iris, aes(x=Sepal.Length, 
                   y=Sepal.Width, 
                   color=Species, 
                   shape=Species)) +
    geom_point(size=6, alpha=0.6)
mytext=paste("Sepal Length = ", iris$Sepal.Length, 
             "\n" , "Sepal Width = ", iris$Sepal.Width, 
             "\n", "Row Number: ",rownames(iris),  sep="")
pp=plotly::plotly_build(p)
style( pp, text=mytext, 
       hoverinfo = "text", 
       traces = c(1, 2, 3) )
```
]
.pull-right[
```{r,  echo=F, fig.align='center', out.height="500px"}
library(plotly)
p=ggplot(iris, aes(x=Sepal.Length, 
                   y=Sepal.Width, 
                   color=Species, 
                   shape=Species)) +
    geom_point(size=6, alpha=0.6)
mytext=paste("Sepal Length = ", iris$Sepal.Length, 
             "\n" , "Sepal Width = ", iris$Sepal.Width, 
             "\n", "Row Number: ",rownames(iris),  sep="")
pp=plotly::plotly_build(p)
style( pp, text=mytext, 
       hoverinfo = "text", 
       traces = c(1, 2, 3) )
```
]
---

# Leaflet

.pull-left[
```{r, eval=F}
library(leaflet)

m <- leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addMarkers(lng=174.768, lat=-36.852, popup="The birthplace of R")
m 
```
]
.pull-right[
```{r, echo=F}
library(leaflet)

m <- leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addMarkers(lng=174.768, lat=-36.852, popup="The birthplace of R")
m 
```
]

---

# Dygraph

.pull-left[
```{r, eval=F}
library(dygraphs)
library(xts)          # To make the convertion data-frame / xts format

# Create data + verify it is date format + change them to xts format:
data=data.frame(time=seq(from=Sys.Date()-40, 
                         to=Sys.Date(), by=1 ), 
                value=runif(41))
str(data$time)
data=xts(x = data$value, order.by = data$time)

# Add points
dygraph(data) %>%
  dyOptions( stemPlot = TRUE)
```
]
.pull-right[
```{r, echo=F, out.height="400px" }
library(dygraphs)
library(xts)          # To make the convertion data-frame / xts format

# Create data + verify it is date format + change them to xts format:
data=data.frame(time=seq(from=Sys.Date()-40, 
                         to=Sys.Date(), by=1 ), 
                value=runif(41))
str(data$time)
data=xts(x = data$value, order.by = data$time)

# Add points
dygraph(data) %>%
  dyOptions( stemPlot = TRUE)
```
]

